Relating two Abstractions for Formal Array Languages

### 2014-04-11 paper.tex
Squeeze some more by using smaller verbatim font
Spelling

Computing with arrays has been a constantly important paradigm in the
history of computing. For this reason we have seen a plethora of
formalisms and languages being developed to aid theoriticians and
developers to produce more suitable abstractions and faster programs.
Examples include Fortran, APL and Matlab but the list goes on and on.
Having many different options to choose from can be a good thing but
it can also lead to a balkanization of the community. When researchers
dedicated to different formalisms can no longer talk to each other,  it
inhibits collaborations and limits the impact of innovation.

@@

As mentioned, one of the most important advantages of Pull arrays is
that they support
fusion\cite{gill1993short,axelsson2010feldspar,keller2010regular}. Fusion
is a program transformation which removes intermediate data structures.
Fusion supercedes transformations such
as loop fusion in optimizing compilers for imperative languages.  Pull
arrays make it particularly simple to achieve fusion. It simply
amounts to inlining the definitions of functions.

@@

There are two different ways of implementing Pull arrays in the
literature. The simplest one comes from using Pull arrays in an
embedded langauge, i.e. when the array language is implemented as a
library in another host language. Examples of such languages are
Pan\cite{elliott2003compiling}, Feldspar\cite{Axelsson:2010:Feldspar},
Obsidian\cite{Svensson:2011:Obsidian} and
Nikola\cite{Mainland:2010:Nikola}. By using the technique of
implementing Pull arrays as \emph{shallow embeddings} on top of a
\emph{deeply embedded} code language, fusion comes completely for free
\cite{svenningsson2013combining}.
In the array library Repa\cite{keller2010regular}, compiler supported
rewrite rules are used to implement fusion \cite{jones2001playing}.

@@

\begin{verbatim}
data Z
data i :. sh

data Shape sh where
  Z :: Shape Z
  (:.) :: Shape tail -> Int -> Shape (tail :. Int)
\end{verbatim}

@@

\begin{verbatim}
data Pull sh a = Pull (Shape sh -> a) (Shape sh)

@@

reverse :: Pull (sh :. Data Length) a ->
           Pull (sh :. Data Length) a
reverse (Pull ixf sh@(_ :. l)) =
 Pull (\(sh' :. i) -> ixf (sh' :. (l - i - 1))) sh
\end{verbatim}

@@

Pull arrays on the other hand were conveived in the context of a
typed functional language. The notation is completely inherited from
the syntax of Haskell \cite{marlow2010haskell}.

@@

The correspondence between MoA and Pull arrays is most clear when
considering Denotaional Normal Form (DNF), \(\psi\)-reductions and
fusion. Both formalisms define operations on arrays by how to index
into the array. In MoA, the indexing function \(\psi\) is used,
whereas for Pull arrays operations are defined as functions from index
to the element at that index. These two forms are equivalent.

@@

Figure \ref{fig:rosetta} gives a translation between function names
used in MoA and Pull arrays. The intention of this figure is to work
as a rosetta stone, helping to translate from one formalism to the
other.

>>>>

Computing with arrays has been a constantly important paradigm in the
history of computing. For this reason we have seen a plethora of
formalisms and languages being developed to aid theoreticians and
developers to produce more suitable abstractions and faster programs.
Examples include FORTRAN, APL and Matlab but the list goes on and on.
Having many different options to choose from can be a good thing but
it can also lead to a balkanization of the community. When researchers
dedicated to different formalisms can no longer talk to each other,  it
inhibits collaborations and limits the impact of innovation.

@@

As mentioned, one of the most important advantages of Pull arrays is
that they support
fusion\cite{gill1993short,axelsson2010feldspar,keller2010regular}. Fusion
is a program transformation which removes intermediate data structures.
Fusion supersedes transformations such
as loop fusion in optimizing compilers for imperative languages.  Pull
arrays make it particularly simple to achieve fusion. It simply
amounts to inlining the definitions of functions.

@@

There are two different ways of implementing Pull arrays in the
literature. The simplest one comes from using Pull arrays in an
embedded language, i.e. when the array language is implemented as a
library in another host language. Examples of such languages are
Pan\cite{elliott2003compiling}, Feldspar\cite{Axelsson:2010:Feldspar},
Obsidian\cite{Svensson:2011:Obsidian} and
Nikola\cite{Mainland:2010:Nikola}. By using the technique of
implementing Pull arrays as \emph{shallow embeddings} on top of a
\emph{deeply embedded} code language, fusion comes completely for free
\cite{svenningsson2013combining}.
In the array library Repa\cite{keller2010regular}, compiler supported
rewrite rules are used to implement fusion \cite{jones2001playing}.

@@

\begin{small}
\begin{Verbatim}
data Z
data i :. sh

data Shape sh where
  Z :: Shape Z
  (:.) :: Shape tail -> Int -> Shape (tail :. Int)
\end{Verbatim}
\end{small}

@@

\begin{small}
\begin{Verbatim}
data Pull sh a = Pull (Shape sh -> a) (Shape sh)

@@

reverse :: Pull (sh :. Data Length) a ->
           Pull (sh :. Data Length) a
reverse (Pull ixf sh@(_ :. l)) =
 Pull (\(sh' :. i) -> ixf (sh' :. (l - i - 1))) sh
\end{Verbatim}
\end{small}

@@

Pull arrays on the other hand were conceived in the context of a
typed functional language. The notation is completely inherited from
the syntax of Haskell \cite{marlow2010haskell}.

@@

The correspondence between MoA and Pull arrays is most clear when
considering Denotational Normal Form (DNF), \(\psi\)-reductions and
fusion. Both formalisms define operations on arrays by how to index
into the array. In MoA, the indexing function \(\psi\) is used,
whereas for Pull arrays operations are defined as functions from index
to the element at that index. These two forms are equivalent.

@@

Figure \ref{fig:rosetta} gives a translation between function names
used in MoA and Pull arrays. The intention of this figure is to work
as a Rosetta stone, helping to translate from one formalism to the
other.
